Improving the portuguese hyphenation rules

### 2023-10-16 article/article.tex
updates

Unfortunately, certain hyphenation rules cannot be implemented using the \TeX{}
hyphenation algorithm. In languages like German, hyphenation can lead to letter
changes or insertions. Additionally, compound words lack hyphens, resulting in
extended letter sequences without visible separation and even repetitions of
the same letter, as seen in examples like \emph{Wasserrinne} and
\emph{Schifffahrt}.  Furthermore, the German spelling reform made some changes,
making it necessary to create a different set of rules for German hyphenation.
For example, the word \emph{Schiffahrt} should be hyphenated as
\emph{Schiff-fahrt}, preserving the \emph{f}s from each word that makes this
compound. The hyphenation should insert a \emph{f} that is not part of the
written form. That was not a problem for the old written form of the word:
\emph{Schifffahrt}. Also, the old hyphenating rules of German grammar stated
the hyphenation \emph{Bäk-ker} for the word \emph{Bäcker}, \emph{Zuk-ker} for
the word \emph{Zucker} and \emph{pak-ken} for the word \emph{packen}. Now those
words are hyphenated as \emph{Bä-cker}, \emph{Zu-cker} and \emph{pa-cken},
respectively. Some words have also changed their hyphenation point after the
spelling reform. For example, \emph{Fen-ster} became \emph{Fens-ter} and
\emph{mei-stens} became \emph{meis-tens}. Some problems in compound word
hyphenation in \TeX{} are discussed in \cite{sojka1995a}.

@@

\section{Patterns for \TeX{} hyphenation}\label{sec-patt4TeX} 
The patterns used in \TeX{} hyphenation are of the form: \texttt{4z1z2}. Each
pattern is made of a sequence of letters and numbers. Odd numbers indicate a
good hyphenation point, whereas even numbers indicate a bad place to break.
The given example states that the sequence has a good breaking point between
the first and the second \emph{z} and an hyphenation should be inhibited before
the first \emph{z} and after the second \emph{z}. For example, the hyphenation
of the word \emph{piz-za}, \emph{fiz-zle} and \emph{mez-zanine} use this rule,
where we see the hyphen placed between the two \emph{z}'s and no hyphen before,
nor after the \emph{z}'s. The pattern may also use period symbol (\emph{.}) to
indicate word boundaries. The pattern \texttt{.s2h2} applies to beginning of
words, implying that the \emph{s} and the \emph{h} should stick together in
beginning of a word and an hyphenation should also be inhibited after the
\emph{h}. For example, this pattern is used in \emph{Sher-lock}.

>>>>

Unfortunately, certain hyphenation rules cannot be implemented using the \TeX{}
hyphenation algorithm. In languages like German, hyphenation can lead to letter
change or insertion. Additionally, compound words lack hyphens, resulting in
extended letter sequences without visible separation and even repetitions of
the same letter, as seen in examples like \emph{Wasserrinne} and
\emph{Schifffahrt}.  Furthermore, the German spelling reform made some changes,
making it necessary to create a different set of rules for German hyphenation.
For example, the word \emph{Schiffahrt} should be hyphenated as
\emph{Schiff-fahrt}, preserving the \emph{f}s from each word that makes this
compound. The hyphenation should insert an \emph{f} that is not part of the
written form. That was not a problem for the old written form of the word:
\emph{Schifffahrt}. Also, the old hyphenating rules of German grammar stated
the hyphenation \emph{Bäk-ker} for the word \emph{Bäcker}, \emph{Zuk-ker} for
the word \emph{Zucker} and \emph{pak-ken} for the word \emph{packen}. Now those
words are hyphenated as \emph{Bä-cker}, \emph{Zu-cker} and \emph{pa-cken},
respectively. Some words have also changed their hyphenation point after the
spelling reform. For example, \emph{Fen-ster} became \emph{Fens-ter} and
\emph{mei-stens} became \emph{meis-tens}. Some problems in compound word
hyphenation in \TeX{} are discussed in \cite{sojka1995a}.

@@

\section{Patterns for \TeX{} hyphenation}\label{sec-patt4TeX} 
If we consider only the Latin alphabet, with no diacritcs, the patterns used 
in \TeX{} hyphenation are of the form: \texttt{\^{}\textbackslash{}.?[0-9]?([a-zA-Z]+[0-9]?)+\textbackslash{}.?\$},
where we have described it using a regular expression\footnote{
    Regular expressions (regex) are powerful search patterns used in text processing to find, match, and manipulate strings of text. They are a fundamental tool in programming and are supported in many programming languages.
    The regex given here uses the Perl syntax and might be broken in the following
    parts: \texttt{\^{}} and \texttt{\$} mark the start and end of the string;
    \texttt{\textbackslash{}.?} specifies an optional period; \texttt{[0-9]?} allows for an optional single digit;
    \texttt{([a-zA-Z]+[0-9]?)+} matches sequences of letters (both uppercase and lowercase) interleaved with optional digits.
}. 
One example of such pattern is \texttt{4z1z2}. Each
pattern is made of a sequence of letters and numbers. Odd numbers indicate a
good hyphenation point, whereas even numbers indicate a bad place to break.
The given example states that the sequence has a good breaking point between
the first and the second \emph{z} and an hyphenation should be inhibited before
the first \emph{z} and after the second \emph{z}. For example, the hyphenation
of the word \emph{piz-za}, \emph{fiz-zle} and \emph{mez-zanine} use this rule,
where we see the hyphen placed between the two \emph{z}'s and no hyphen before,
nor after the \emph{z}'s. The pattern may also use period symbol (\emph{.}) to
indicate word boundaries. The pattern \texttt{.s2h2} applies to beginning of
words, implying that the \emph{s} and the \emph{h} should stick together in
beginning of a word and an hyphenation should also be inhibited after the
\emph{h}. For example, this pattern is used in \emph{Sher-lock}.
